---
title: "Bayesian Project"
author: "Satoshi Koiso, Zaw Wai Yan Bo, Pyone Yadanar Paing"
date: "2026-05-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r set up}
## Load libraries 
library(haven)
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(rjags)
# library(rethinking)
library(coda)

set.seed(123)
```

Data Preparation and cleaning
Datasets are downloaded from website below. 
https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2021-2023


```{r Data Cleaning}
## Read datasets
a1c <- read_xpt("GHB_L.xpt")
demo <- read_xpt("DEMO_L.xpt")

# merge datasets
nhanes <- merge(a1c, demo, by = "SEQN", all = TRUE)
d <- setDT(nhanes)

# Limit the dataset to 12+ years old because HbA1c is measured 12+ years and complete obs of A1c
d <- d[RIDAGEYR >= 12 & !is.na(LBXGH)]
summary(d$LBXGH)

# Create dummy variable for male
d$male <- ifelse(d$RIAGENDR == 1, 1, 0)
```

Prior predictive check
We will apply Bayesian multiple linear regression to predict posterior distribution of HbA1c and simulate results using the posterior predictive distribution. Outcome variable would be HbA1c as continuous variable (%) and predictor variables would be age as continuous variable (years) and sex as categorical variable. We will center age ($x_i$) and create dummy variable for sex: male ($m_i$). We will also include interaction term between age (years) and $m_i$.

Linear Model we will fit: 

$$\mu_i = \beta_0 + \beta_1 (x_i -\bar{x}) + \beta_2 m_i + \beta_3 (m_i (x_i -\bar{x}))$$


Prior: Our prior for average HbA1c level in the population is below 5.7%. 
- $\beta_0 \sim \mathcal{Normal}(5.5,1)$: Intercept for women
- $\beta_1 \sim \mathcal{Normal}(0,0.05)$: Women’s slope
- $\beta_2 \sim \mathcal{Normal}(0,0.05)$: Intercept difference
- $\beta_3 \sim \mathcal{Normal}(0,0.05)$: Slope difference

```{r Prior predictive check}
n_obs <- nrow(d)
xbar <- mean(d$RIDAGEYR)
x_centered <- d$RIDAGEYR - xbar
x_range <- range(x_centered)


# Number of simulations
nsim <- 200

# 1. Simulate parameters from our priors
beta0_prior <- rnorm(nsim, mean = 5.5, sd = 1)
beta1_prior <- rnorm(nsim, mean = 0, sd = 0.05)
beta2_prior <- rnorm(nsim, mean = 0, sd = 0.05)
beta3_prior <- rnorm(nsim, mean = 0, sd = 0.05)
sigma_prior <- runif(nsim, min = 0, max = 0.1) # we have not discussed sigma yet

# 3. Simulate full datasets from the prior (one dataset per draw)
ynew <- matrix(nrow = n_obs, ncol = nsim)

for(j in 1:nsim) {
  ynew[, j] <- rnorm(n_obs, mean = beta0_prior[j] + beta1_prior[j] * x_centered + beta2_prior[j]* d$male + beta3_prior[j]*d$male*x_centered, sd = sigma_prior[j])
}

ynew_df <- as.data.frame(ynew)
ynew_df$id <- 1:nrow(ynew_df)
ynew_long <- reshape(ynew_df, direction = "long", idvar = "id",
                     varying = list(1:nsim), timevar = "simulation",
                     v.names = "HbA1c", times = 1:nsim)
rownames(ynew_long) <- NULL


# 3. Visualize the Prior Predictive Distribution
# Plot the predictive check
plot(NULL, xlim = x_range, ylim = c(0,20),
     xlab = "Centered Age", ylab = "HbA1C",
     main = "Implied Prior Regression Lines")

# for (i in 1:nsim){
#   curve(beta0_prior[i] + beta1_prior[i] * x + beta2_prior[i]* d$male + beta3_prior[i]*d$male*x,
#         from = x_range[1], to = x_range[2],
#         add = TRUE, col = adjustcolor("black", alpha.f = 0.2))
#   
# }


ggplot() +
  # 200 simulated datasets
  geom_line(data = ynew_long, aes(x = HbA1c, group = simulation), stat = "density", color = "lightblue", alpha = 0.9) +
  # Actual observed data
  geom_density(data = d, aes(x = LBXGH), color = "black", linewidth = 1.2) +
  coord_cartesian(ylim = c(0, 0.7))+
  theme_minimal() +
  labs(title = "Prior Predictive Check: Intercept-Only Model",
       subtitle = "Black line = Actual Data | Blue lines = Simulated Datasets",
       x = "HbA1c", y = "Density")
```

JAGS model 
```{r JAGS model}
model_code_interaction <- "model{
  # --- LIKELIHOOD ---
  for(i in 1:n){
    y[i] ~ dnorm(mu[i], tau) 
    
    # Linear predictor with interaction
    mu[i] <- beta0 + beta1 * (x[i] - xbar) + beta2 * m[i] + beta3 * m[i] * (x[i] - xbar)
    
    # Posterior Predictive
    ynew[i] ~ dnorm(mu[i], tau)
  }
 
  # --- PRIORS ---
  sigma ~ dunif(0, 50)
  tau    <- pow(sigma, -2)
  
  # Note: JAGS dnorm uses Precisions = 1/Variance
  beta0 ~ dnorm(5.5, pow(1, -2))
  beta1 ~ dnorm(0, pow(0.05, -2))
  beta2 ~ dnorm(0, pow(0.05, -2))
  beta3 ~ dnorm(0, pow(0.05, -2))
  
  # Generated Quantities: Calculate Men-specific parameters directly inside JAGS
  beta0_m <- beta0 + beta2
  beta1_m <- beta1 + beta3
}"

data_list_int <- list(
  y = d$LBXGH, 
  x = d$RIDAGEYR, 
  m = d$male, 
  xbar = xbar, 
  n = n_obs
)

m4 <- jags.model(
  textConnection(model_code_interaction),
  data = data_list_int,
  n.chains = 3,
  n.adapt = 1000,
  quiet = TRUE
)

# Burn-in
update(m4, n.iter = 2000)

# Draw posterior samples
samples_int <- coda.samples(
  model = m4,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "beta0_m", "beta1_m", "sigma"),
  n.iter = 10000
)

samps_int_df <- as.data.frame(samples_int[[1]])

summary(samples_int)
posterior_summary <- data.frame(
  mean = apply(samps_int_df, 2, mean),
  sd = apply(samps_int_df, 2, sd),
  q2.5 = apply(samps_int_df, 2, quantile, probs = 0.025),
  q97.5 = apply(samps_int_df, 2, quantile, probs = 0.975)
)

round(posterior_summary, 2)

```

MCMC Diagnostics 
```{r MCMC Diagnositcs}
# Trace plots
plot(samples_int, density = FALSE, main = "")

# Gelman-Rubin diagnostic: R-hat
gelman.diag(samples_int)$psrf

# Effective sample size
effectiveSize(samples_int)

# Optional: autocorrelation plots
autocorr.plot(samples_int)
```

Posterior predictive check

After confirming that the MCMC chains converged, we conducted posterior predictive checks to evaluate whether the fitted Bayesian regression model can reproduce the main features of the observed HbA1c distribution. Following the lecture workflow, we simulated replicated HbA1c values from the posterior predictive distribution and compared them with the observed NHANES HbA1c values.

```{r Posterior predictive check}
ynew[i] ~ dnorm(mu[i], tau)

# Draw posterior predictive samples from the fitted JAGS model
samples_pred <- jags.samples(
  model = m4,
  variable.names = c("ynew", "mu"),
  n.iter = 1000
)

# Check dimensions
dim(samples_pred$ynew)

# Extract posterior predictive array
ynew_array <- samples_pred$ynew

# Combine iterations across chains
# Result: rows = individuals, columns = posterior predictive draws
ynew_mat <- do.call(
  cbind,
  lapply(1:dim(ynew_array)[3], function(ch) ynew_array[, , ch])
)

# Extract posterior mean array if needed later
mu_array <- samples_pred$mu
mu_mat <- do.call(
  cbind,
  lapply(1:dim(mu_array)[3], function(ch) mu_array[, , ch])
)

# Check final dimensions
dim(ynew_mat)
dim(mu_mat)

## 2. Density overlay PPC
set.seed(123)

# Number of posterior predictive datasets to show
n_show <- 100
idx <- sample(1:ncol(ynew_mat), n_show)

# Observed HbA1c density
plot(
  density(d$LBXGH),
  col = "black",
  lwd = 2,
  xlab = "HbA1c (%)",
  ylab = "Density",
  main = "Posterior Predictive Check: HbA1c Distribution",
  xlim = c(3, 18),
  ylim = c(0, 1)
)

# Overlay posterior predictive densities
for (s in idx) {
  y_rep <- ynew_mat[, s]
  y_rep <- y_rep[is.finite(y_rep)]
  
  # Defensive coding: skip if no variation
  if (length(unique(y_rep)) < 2) next
  
  dens <- tryCatch(
    density(y_rep),
    error = function(e) NULL
  )
  
  if (!is.null(dens)) {
    lines(
      dens,
      col = adjustcolor("skyblue", alpha.f = 0.12),
      lwd = 1
    )
  }
}

# Add observed density again on top
lines(density(d$LBXGH), col = "black", lwd = 2)

legend(
  "topright",
  legend = c("Observed HbA1c", "Posterior predictive draws"),
  col = c("black", adjustcolor("skyblue", alpha.f = 0.6)),
  lwd = c(2, 1),
  bty = "n"
)

# Observed summary statistics
obs_mean <- mean(d$LBXGH)
obs_sd <- sd(d$LBXGH)
obs_above57 <- mean(d$LBXGH >= 5.7)
obs_above65 <- mean(d$LBXGH >= 6.5)

# Posterior predictive summary statistics
rep_means <- apply(ynew_mat, 2, mean)
rep_sds <- apply(ynew_mat, 2, sd)
rep_above57 <- apply(ynew_mat, 2, function(x) mean(x >= 5.7))
rep_above65 <- apply(ynew_mat, 2, function(x) mean(x >= 6.5))

# Create summary table
ppc_table <- data.frame(
  Statistic = c(
    "Mean HbA1c",
    "SD HbA1c",
    "Proportion HbA1c >= 5.7%",
    "Proportion HbA1c >= 6.5%"
  ),
  Observed = c(
    obs_mean,
    obs_sd,
    obs_above57,
    obs_above65
  ),
  PPC_lower_2.5 = c(
    quantile(rep_means, 0.025),
    quantile(rep_sds, 0.025),
    quantile(rep_above57, 0.025),
    quantile(rep_above65, 0.025)
  ),
  PPC_median = c(
    quantile(rep_means, 0.5),
    quantile(rep_sds, 0.5),
    quantile(rep_above57, 0.5),
    quantile(rep_above65, 0.5)
  ),
  PPC_upper_97.5 = c(
    quantile(rep_means, 0.975),
    quantile(rep_sds, 0.975),
    quantile(rep_above57, 0.975),
    quantile(rep_above65, 0.975)
  )
)

ppc_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

# Check whether posterior predictive draws produce impossible HbA1c values
rep_negative <- apply(ynew_mat, 2, function(x) mean(x < 0))
rep_extreme_high <- apply(ynew_mat, 2, function(x) mean(x > 18))

cat("Average posterior predictive proportion with HbA1c < 0:",
    round(mean(rep_negative), 5), "\n")

cat("Average posterior predictive proportion with HbA1c > 18:",
    round(mean(rep_extreme_high), 5), "\n")
```
