---
title: "Bayesian Project"
author: "Satoshi Koiso, Zaw Wai Yan Bo, Pyone Yadanar Paing"
date: "2026-05-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r set up}
## Load libraries 
library(haven)
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(rjags)
library(rethinking)
library(coda)
```

Data Preparation and cleaning
Datasets are downloaded from website below. 
https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2021-2023


```{r Data Cleaning}
## Read datasets
a1c <- read_xpt("GHB_L.xpt")
demo <- read_xpt("DEMO_L.xpt")

# merge datasets
nhanes <- merge(a1c, demo, by = "SEQN", all = TRUE)
d <- setDT(nhanes)

# Limit the dataset to 12+ years old because HbA1c is measured 12+ years and complete obs of A1c
d <- d[RIDAGEYR >= 12 & !is.na(LBXGH)]
summary(d$LBXGH)

# Create dummy variable for male
d$male <- ifelse(d$RIAGENDR == 1, 1, 0)
```

Prior predictive check
We will apply Bayesian multiple linear regression to predict posterior distribution of HbA1c and simulate results using the posterior predictive distribution. Outcome variable would be HbA1c as continuous variable (%) and predictor variables would be age as continuous variable (years) and sex as categorical variable. We will center age ($x_i$) and create dummy variable for sex: male ($m_i$). We will also include interaction term between age (years) and $m_i$.

Linear Model we will fit: 

$$\mu_i = \beta_0 + \beta_1 (x_i -\bar{x}) + \beta_2 m_i + \beta_3 (m_i (x_i -\bar{x}))$$


Prior: Our prior for average HbA1c level in the population is below 5.7%. 
- $\beta_0 \sim \mathcal{Normal}(5.5,1)$: Intercept for women
- $\beta_1 \sim \mathcal{Normal}(0,0.05)$: Women’s slope
- $\beta_2 \sim \mathcal{Normal}(0,0.05)$: Intercept difference
- $\beta_3 \sim \mathcal{Normal}(0,0.05)$: Slope difference

```{r Prior predictive check}
xbar <- mean(d$LBXGH)
x_centered <- d$LBXGH - xbar
x_range <- range(x_centered)
```

JAGS model 
```{r JAGS model}
model_code_interaction <- "model{
  # --- LIKELIHOOD ---
  for(i in 1:n){
    y[i] ~ dnorm(mu[i], tau) 
    
    # Linear predictor with interaction
    mu[i] <- beta0 + beta1 * (x[i] - xbar) + beta2 * m[i] + beta3 * m[i] * (x[i] - xbar)
    
    # Posterior Predictive
    ynew[i] ~ dnorm(mu[i], tau)
  }
 
  # --- PRIORS ---
  sigma ~ dunif(0, 50)
  tau    <- pow(sigma, -2)
  
  # Note: JAGS dnorm uses Precisions = 1/Variance
  beta0 ~ dnorm(5.5, pow(1, -2))
  beta1 ~ dnorm(0, pow(0.05, -2))
  beta2 ~ dnorm(0, pow(0.05, -2))
  beta3 ~ dnorm(0, pow(0.05, -2))
  
  # Generated Quantities: Calculate Men-specific parameters directly inside JAGS
  beta0_m <- beta0 + beta2
  beta1_m <- beta1 + beta3
}"

data_list_int <- list(
  y = d$LBXGH, 
  x = d$RIDAGEYR, 
  m = d$male, 
  xbar = xbar, 
  n = n_obs
)

m4 <- jags.model(textConnection(model_code_interaction), data = data_list_int, quiet = TRUE)

samples_int <- coda.samples(
  m4, 
  variable.names = c("beta0", "beta1", "beta2", "beta3", "beta0_m", "beta1_m", "sigma"), 
  n.iter = 10000
)

samps_int_df <- as.data.frame(samples_int[[1]])

# Display summary of all beta coefficients
precis(samps_int_df, prob = 0.95, digits = 2)


```


Posterior predictive check
```{r Posterior predictive check}

```

MCMC Diagnostics 
```{r MCMC Diagnositcs}

```
