---
title: "Bayesian Project"
author: "Satoshi Koiso, Zaw Wai Yan Bo, Pyone Yadanar Paing"
date: "2026-05-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r set up}
## Load libraries 
library(haven)
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(rjags)
# library(rethinking)
library(coda)

set.seed(123)
```

Data Preparation and cleaning
Datasets are downloaded from website below. 
https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2021-2023


```{r Data Cleaning}
## Read datasets
a1c <- read_xpt("GHB_L.xpt")
demo <- read_xpt("DEMO_L.xpt")
diq <- read_xpt("DIQ_L.xpt")

# merge datasets
nhanes <- merge(a1c, demo, by = "SEQN", all = TRUE)
nhanes <- merge(nhanes, diq, by = "SEQN", all = TRUE)

d <- setDT(nhanes)

# Limit the dataset to 12+ years old because HbA1c is measured 12+ years and complete obs of A1c
d <- d[RIDAGEYR >= 12 & !is.na(LBXGH)]
summary(d$LBXGH)

# Create dummy variable for male
d$male <- ifelse(d$RIAGENDR == 1, 1, 0)

```

Prior predictive check
We will apply Bayesian multiple linear regression to predict posterior distribution of HbA1c and simulate results using the posterior predictive distribution. Outcome variable would be HbA1c as continuous variable (%) and predictor variables would be age as continuous variable (years) and sex as categorical variable. We will center age ($x_i$) and create dummy variable for sex: male ($m_i$). We will also include interaction term between age (years) and $m_i$.

Linear Model we will fit: 

$$\mu_i = \beta_0 + \beta_1 (x_i -\bar{x}) + \beta_2 m_i + \beta_3 (m_i (x_i -\bar{x}))$$


Prior: Our prior for average HbA1c level in the population is below 5.7%. 
- $\beta_0 \sim \mathcal{Normal}(5.5,0.5)$: Intercept for women
- $\beta_1 \sim \mathcal{Normal}(0.01,0.05)$: Women’s slope
- $\beta_2 \sim \mathcal{Normal}(0.11,0.05)$: Intercept difference
- $\beta_3 \sim \mathcal{Normal}(0,0.05)$: Slope difference

```{r Prior predictive check}
# ── Setup ──────────────────────────────────────────────────────────────
n_obs    <- nrow(d)
xbar     <- mean(d$RIDAGEYR)
x_centered <- d$RIDAGEYR - xbar

# ── Prior predictive check (log-normal model with elevated covariate) ──
nsim <- 200

# Priors on LOG scale
# Check log scale first
cat("mean(log(HbA1c)):", mean(log(d$LBXGH)), "\n")
cat("sd(log(HbA1c)):",   sd(log(d$LBXGH)),   "\n")

# ── Informed priors from data diagnostics ──────────────────────────────
# mean(log(HbA1c)) normal group  = 1.656
# mean(log(HbA1c)) elevated group = 1.865
# sd(log(HbA1c)) overall          = 0.149
# So sigma on log scale ~ 0.10-0.15 within groups

nsim <- 200

beta0_prior <- rnorm(nsim, mean = 1.7, sd = 0.05)   # log-normal group intercept
beta1_prior <- rnorm(nsim, mean = 0.003, sd = 0.003)   # age slope (log scale)
beta2_prior <- rnorm(nsim, mean = 0.01,  sd = 0.005)   # sex offset (log scale)
beta3_prior <- rnorm(nsim, mean = 0,     sd = 0.003)   # interaction
sigma_prior <- runif(nsim, min = 0, max = 0.1)   # sd of hb1ac - 0.6 

ynew_log <- matrix(nrow = n_obs, ncol = nsim)

for(j in 1:nsim) {
  ynew_log[, j] <- exp(rnorm(n_obs, 
          mean = beta0_prior[j] +
          beta1_prior[j] * x_centered +
          beta2_prior[j] * d$male +
          beta3_prior[j] * d$male * x_centered, 
          sd = sigma_prior[j]))
}


# ── Reshape for ggplot ─────────────────────────────────────────────────
ynew_df <- as.data.frame(ynew_log)
ynew_df$id <- 1:nrow(ynew_df)
ynew_long <- reshape(ynew_df,
                     direction = "long",
                     idvar     = "id",
                     varying   = list(1:nsim),
                     timevar   = "simulation",
                     v.names   = "HbA1c",
                     times     = 1:nsim)
rownames(ynew_long) <- NULL

# ── Plot ───────────────────────────────────────────────────────────────
ggplot() +
  geom_line(
    data  = ynew_long,
    aes(x = HbA1c, group = simulation),
    stat  = "density",
    color = "lightblue",
    alpha = 0.5
  ) +
  geom_density(
    data      = d,
    aes(x     = LBXGH),
    color     = "black",
    linewidth = 1.5
  ) +
  coord_cartesian(xlim = c(3, 15), ylim = c(0, 1.5)) +
  theme_minimal() +
  labs(
    title    = "Prior Predictive Check: Log-Normal Model",
    subtitle = "Black line = Actual Data | Blue lines = Simulated Datasets",
    x        = "HbA1c (%)",
    y        = "Density"
  )



```

JAGS model 
```{r JAGS model}
model_code_interaction <- "model{
  # --- LIKELIHOOD ---
  for(i in 1:n){
    y[i] ~ dnorm(mu[i], tau) 
    
    # Linear predictor with interaction
    mu[i] <- beta0 + beta1 * (x[i] - xbar) + beta2 * m[i] + beta3 * m[i] * (x[i] - xbar)
    
    # Posterior Predictive
    ynew[i] ~ dnorm(mu[i], tau)
  }
 
  # --- PRIORS ---
  sigma ~ dunif(0, 0.1)
  tau    <- pow(sigma, -2)
  
  # Note: JAGS dnorm uses Precisions = 1/Variance
  beta0 ~ dnorm(1.7, pow(0.05, -2))
  beta1 ~ dnorm(0.003, pow(0.003, -2))
  beta2 ~ dnorm(0.01, pow(0.005, -2))
  beta3 ~ dnorm(0, pow(0.003, -2))
  
  # Generated Quantities: Calculate Men-specific parameters directly inside JAGS
  beta0_m <- beta0 + beta2
  beta1_m <- beta1 + beta3
}"

data_list_int <- list(
  y = log(d$LBXGH), 
  x = d$RIDAGEYR, 
  m = d$male, 
  xbar = xbar, 
  n = n_obs
)

m4 <- jags.model(textConnection(model_code_interaction), 
                 data = data_list_int, 
                 n.chains = 3,
                 quiet = TRUE)

# Burn-in
update(m4, n.iter = 2000)

samples_int <- coda.samples(
  m4,
  variable.names = c("beta0", "beta1", "beta2", "beta3", 
                     "beta0_m", "beta1_m", "sigma"),
  n.iter = 10000
)

samps_int_df <- as.data.frame(samples_int[[1]])

beta_cols  <- grep("^beta",  names(samps_int_df), value = TRUE)
sigma_cols <- grep("^sigma", names(samps_int_df), value = TRUE)

exp_hpd   <- HPDinterval(as.mcmc(exp(samps_int_df[, beta_cols])), prob = 0.95)
sigma_hpd <- HPDinterval(as.mcmc(samps_int_df[, sigma_cols, drop = FALSE]), prob = 0.95)

exp_summary <- data.frame(
  parameter = beta_cols,
  exp_mean  = round(colMeans(exp(samps_int_df[, beta_cols])), 4),
  exp_lo    = round(exp_hpd[, "lower"], 4),
  exp_hi    = round(exp_hpd[, "upper"], 4)
)

sigma_summary <- data.frame(
  parameter = sigma_cols,
  exp_mean  = round(colMeans(samps_int_df[, sigma_cols, drop = FALSE]), 4),
  exp_lo    = round(sigma_hpd[, "lower"], 4),
  exp_hi    = round(sigma_hpd[, "upper"], 4)
)

full_summary <- rbind(exp_summary, sigma_summary)
print(full_summary)

# Bias correction term 
bias <- 0.5 * samps_int_df$sigma^2

corrected <- data.frame(
  beta0_corrected   = exp(samps_int_df$beta0   + bias),
  beta0_m_corrected = exp(samps_int_df$beta0_m + bias),
  beta1_corrected   = exp(samps_int_df$beta1   + bias),
  beta1_m_corrected = exp(samps_int_df$beta1_m + bias),
  beta2_corrected   = exp(samps_int_df$beta2   + bias),
  beta3_corrected   = exp(samps_int_df$beta3   + bias),
  sigma             = samps_int_df$sigma
)

# HPD intervals 
hpd <- HPDinterval(as.mcmc(corrected), prob = 0.95)

# Equal-tail intervals 
ci_lo <- apply(corrected, 2, quantile, 0.025)
ci_hi <- apply(corrected, 2, quantile, 0.975)

# Combined summary table 
corrected_summary <- data.frame(
  parameter = rownames(hpd),
  post_mean = round(colMeans(corrected), 4),
  # HPD
  hpd_lower = round(hpd[, "lower"], 4),
  hpd_upper = round(hpd[, "upper"], 4),
  # Equal-tail
  ci_lower  = round(ci_lo, 4),
  ci_upper  = round(ci_hi, 4)
)

print(corrected_summary)
```


Posterior predictive check
```{r Posterior predictive check}

# Draw posterior predictive samples from the fitted JAGS model
samples_pred <- jags.samples(
  model = m4,
  variable.names = c("ynew", "mu"),
  n.iter = 1000
)

# Check dimensions
dim(samples_pred$ynew)

# Extract posterior predictive array
ynew_array <- samples_pred$ynew

# Combine iterations across chains
ynew_mat <- do.call(
  cbind,
  lapply(1:dim(ynew_array)[3], function(ch) ynew_array[, , ch])
)

# Extract posterior mean array if needed later
mu_array <- samples_pred$mu
mu_mat <- do.call(
  cbind,
  lapply(1:dim(mu_array)[3], function(ch) mu_array[, , ch])
)

# Check final dimensions
dim(ynew_mat)
dim(mu_mat)

ynew_mat_original <- exp(ynew_mat) 


## 2. Density overlay PPC   ## Lecture 12
set.seed(123)

# Number of posterior predictive datasets to show
n_show <- 100
idx <- sample(1:ncol(ynew_mat_original), n_show)

# Observed HbA1c density
plot(
  density(d$LBXGH),
  col = "black",
  lwd = 2,
  xlab = "HbA1c (%)",
  ylab = "Density",
  main = "Posterior Predictive Check: HbA1c Distribution",
  xlim = c(3, 18),
  ylim = c(0, 1)
)

# Overlay posterior predictive densities
for (s in idx) {
  y_rep <- ynew_mat_original[, s]
  y_rep <- y_rep[is.finite(y_rep)]
 
  # Defensive coding: skip if no variation
  if (length(unique(y_rep)) < 2) next
 
  dens <- tryCatch(
    density(y_rep),
    error = function(e) NULL
  )
 
  if (!is.null(dens)) {
    lines(
      dens,
      col = adjustcolor("skyblue", alpha.f = 0.12),
      lwd = 1
    )
  }
}

# Add observed density
lines(density(d$LBXGH), col = "black", lwd = 2)

legend(
  "topright",
  legend = c("Observed HbA1c", "Posterior predictive draws"),
  col = c("black", adjustcolor("skyblue", alpha.f = 0.6)),
  lwd = c(2, 1),
  bty = "n"
)

# Observed summary statistics
obs_mean <- mean(d$LBXGH)
obs_sd <- sd(d$LBXGH)
obs_above57 <- mean(d$LBXGH >= 5.7)
obs_above65 <- mean(d$LBXGH >= 6.5)

# Posterior predictive summary statistics
rep_means <- apply(ynew_mat_original, 2, mean)
rep_sds <- apply(ynew_mat_original, 2, sd)
rep_above57 <- apply(ynew_mat_original, 2, function(x) mean(x >= 5.7))
rep_above65 <- apply(ynew_mat_original, 2, function(x) mean(x >= 6.5))

# Create summary table
ppc_table <- data.frame(
  Statistic = c(
    "Mean HbA1c",
    "SD HbA1c",
    "Proportion HbA1c >= 5.7%",
    "Proportion HbA1c >= 6.5%"
  ),
  Observed = c(
    obs_mean,
    obs_sd,
    obs_above57,
    obs_above65
  ),
  PPC_lower_2.5 = c(
    quantile(rep_means, 0.025),
    quantile(rep_sds, 0.025),
    quantile(rep_above57, 0.025),
    quantile(rep_above65, 0.025)
  ),
  PPC_median = c(
    quantile(rep_means, 0.5),
    quantile(rep_sds, 0.5),
    quantile(rep_above57, 0.5),
    quantile(rep_above65, 0.5)
  ),
  PPC_upper_97.5 = c(
    quantile(rep_means, 0.975),
    quantile(rep_sds, 0.975),
    quantile(rep_above57, 0.975),
    quantile(rep_above65, 0.975)
  )
)

ppc_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

# Check posterior predictive draws for HbA1c values
rep_negative <- apply(ynew_mat_original, 2, function(x) mean(x < 0))
rep_extreme_high <- apply(ynew_mat_original, 2, function(x) mean(x > 18))

cat("Average posterior predictive proportion with HbA1c < 0:",
    round(mean(rep_negative), 5), "\n")

cat("Average posterior predictive proportion with HbA1c > 18:",
    round(mean(rep_extreme_high), 5), "\n")
```

MCMC Diagnostics 
```{r MCMC Diagnositcs}
# Trace plots
plot(samples_int, density = FALSE, main = "")

# Gelman-Rubin diagnostic: R-hat
gelman.diag(samples_int[, c("beta0", "beta1", "beta2", "beta3", "sigma")])$psrf

# Effective sample size
effectiveSize(samples_int)

# Optional: autocorrelation plots
autocorr.plot(samples_int)
```
